---
title: "Chapter 6"
output: html_document
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Chapter 6

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
library(survival)
library(tidyr)
library(dplyr)
library(knitr)
```

## Model Selection and Interpretation 

## 6.1 - Example of genetics confounding data

## 6.2 Categorical and Continuous Covariates 

```{r cars}

race <- factor(c("black", "black", "white", "white", "other",  "other"))  
age <- c(48, 52, 87, 82, 67, 53)
model.matrix(~ race + age)[,-1] 

race <- relevel(race, ref="white")
model.matrix(~ race + age)[,-1] 

```

In this example we have three covariates, say, z1, z2, and z3, the first two of  which are dummy variables for black race and other race, and the third a continuous  variable, age.


```{r}

# Interaction between age and race, 
# The interaction terms (last two columns) are just the product of the first two columns  and the third (age) column.  

model.matrix(~ race + age + race:age)[,-1]

```

### Example

```{r}

age <- runif(n=60, min=40, max=80) 

race <- factor(c(rep("white", 20), rep("black", 20),  rep("other", 20))) 
race <- relevel(race, ref="white")  

log.rate.vec <- -4.5 + c(rep(0,20), rep(1,20), rep(2,20))  + age*0.05 

tt <- rexp(n=60, rate=exp(log.rate.vec))  
status <- rep(1, 60)  
library(survival) 
result.cox <- coxph(Surv(tt, status) ~ race + age) 
summary(result.cox) 

# 1.02, 1.72 and 0.05 are log hazard ratios.
# 2.78, 5.61 and 1.05 are exponentiated. Meaning 2.78 higher risk for black.

```

## 6.3 Hypothesis Testing for Nested Models 

```{r}

data(pharmacoSmoking)

attach(pharmacoSmoking)

pharmacoSmoking %>% head(25) %>%  kable()

survfit(Surv(pharmacoSmoking$ttr,pharmacoSmoking$relapse) ~ pharmacoSmoking$grp)

coxph(Surv(pharmacoSmoking$ttr,pharmacoSmoking$relapse) ~ pharmacoSmoking$ageGroup4)


modelA.coxph <- coxph(Surv(ttr, relapse) ~ ageGroup4)
modelB.coxph <- coxph(Surv(ttr, relapse) ~ employment) 
modelC.coxph <- coxph(Surv(ttr, relapse) ~ ageGroup4 + employment)

modelA.coxph
modelB.coxph
modelC.coxph # ageGroup450-64, 

# term “ageGroup4” or the term “employment”  belong in the model? 

logLik(modelA.coxph)  
logLik(modelB.coxph) 
logLik(modelC.coxph) 

2*( logLik(modelC.coxph) -logLik(modelA.coxph) )
pchisq(2*( logLik(modelC.coxph) -logLik(modelA.coxph) ), 
       df=2, 
       lower.tail=F) # nope

# term “ageGroup4” or the term “employment”  belong in the model? 

logLik(modelA.coxph)  

logLik(modelC.coxph) 

2*( logLik(modelC.coxph) -logLik(modelB.coxph) )
pchisq(2*( logLik(modelC.coxph) - logLik(modelB.coxph)  ), 
       df=3, 
       lower.tail=F) # nope

# We thus conclude that “employment” belongs in the model if “ageGroup4” is also  included, since the p-value for the former is extremely small


```

```{r}

# ANOVA
anova(modelA.coxph, modelC.coxph)
```
```{r AIC}

AIC(modelA.coxph)
AIC(modelB.coxph)
AIC(modelC.coxph) # <- this one

```

### Step AIC
```{r}

modelAll.coxph <- coxph(Surv(pharmacoSmoking$ttr, pharmacoSmoking$relapse) ~ pharmacoSmoking$grp + pharmacoSmoking$gender + pharmacoSmoking$race + pharmacoSmoking$employment + pharmacoSmoking$yearsSmoking + pharmacoSmoking$levelSmoking +  pharmacoSmoking$ageGroup4 + pharmacoSmoking$priorAttempts + pharmacoSmoking$longestNoSmoke)  

result.step <- step(modelAll.coxph, scope=list(upper=~ pharmacoSmoking$grp +  pharmacoSmoking$gender + pharmacoSmoking$race + pharmacoSmoking$employment + pharmacoSmoking$yearsSmoking +  pharmacoSmoking$levelSmoking + pharmacoSmoking$ageGroup4 + pharmacoSmoking$priorAttempts +  pharmacoSmoking$longestNoSmoke, lower=~pharmacoSmoking$grp))  

result.step

```

## 6.5 Smoothing

> A classical treatment of splines is de Boor  [13], and their use in statistics has been discussed by many authors. In survival  analysis, an effective method of finding a smoothing spline is via “penalized partial  likelihood.”

```{r}

modelS4.coxph <- coxph(Surv(pharmacoSmoking$ttr, pharmacoSmoking$relapse) ~ pharmacoSmoking$grp + pharmacoSmoking$employment +  + pspline(pharmacoSmoking$age, df=4))  
modelS4.coxph 

# termplot(modelS4.coxph, se=T, terms=3, ylabs="Log hazard")


```












