---
title: "Chapter 6"
output: html_document
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Chapter 6

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
library(survival)
library(tidyr)
library(dplyr)
library(knitr)
library(asaur)
library(survminer)
```

## Model Selection and Interpretation 

## 6.1 - Example of genetics confounding data

## 6.2 Categorical and Continuous Covariates 

```{r cars}

race <- factor(c("black", "black", "white", "white", "other",  "other"))  
age <- c(48, 52, 87, 82, 67, 53)
model.matrix(~ race + age)[,-1] 

race <- relevel(race, ref="white")
model.matrix(~ race + age)[,-1] 

```

In this example we have three covariates, say, z1, z2, and z3, the first two of  which are dummy variables for black race and other race, and the third a continuous  variable, age.


```{r}

# Interaction between age and race, 
# The interaction terms (last two columns) are just the product of the first two columns  and the third (age) column.  

model.matrix(~ race + age + race:age)[,-1]

```

### Example

```{r}

age <- runif(n=60, min=40, max=80) 

race <- factor(c(rep("white", 20), rep("black", 20),  rep("other", 20))) 
race <- relevel(race, ref="white")  

log.rate.vec <- -4.5 + c(rep(0,20), rep(1,20), rep(2,20))  + age*0.05 

tt <- rexp(n=60, rate=exp(log.rate.vec))  
status <- rep(1, 60)  
library(survival) 
result.cox <- coxph(Surv(tt, status) ~ race + age) 
summary(result.cox) 

# 1.02, 1.72 and 0.05 are log hazard ratios.
# 2.78, 5.61 and 1.05 are exponentiated. Meaning 2.78 higher risk for black.

```

## 6.3 Hypothesis Testing for Nested Models 

When constructing statistical tests, it is necessary to compare what are  called `“nested” models`. 

That is, when comparing two models, the covariates of one  model must be a subset of the covariates in the other.



```{r}

attach(pharmacoSmoking)
data(pharmacoSmoking)


pharmacoSmoking %>% head(25) %>%  kable()

survfit(Surv(pharmacoSmoking$ttr,pharmacoSmoking$relapse) ~ pharmacoSmoking$grp)

coxph(Surv(pharmacoSmoking$ttr,pharmacoSmoking$relapse) ~ pharmacoSmoking$ageGroup4)


modelA.coxph <- coxph(Surv(ttr, relapse) ~ ageGroup4)
modelB.coxph <- coxph(Surv(ttr, relapse) ~ employment) 
modelC.coxph <- coxph(Surv(ttr, relapse) ~ ageGroup4 + employment)

modelA.coxph
modelB.coxph
modelC.coxph # ageGroup450-64, 

# term “ageGroup4” or the term “employment”  belong in the model? 

logLik(modelA.coxph)  
logLik(modelB.coxph) 
logLik(modelC.coxph) 

2*( logLik(modelC.coxph) -logLik(modelA.coxph) )
pchisq(2*( logLik(modelC.coxph) -logLik(modelA.coxph) ), 
       df=2, 
       lower.tail=F) # nope

# term “ageGroup4” or the term “employment”  belong in the model? 

logLik(modelA.coxph)  

logLik(modelC.coxph) 

2*( logLik(modelC.coxph) -logLik(modelB.coxph) )
pchisq(2*( logLik(modelC.coxph) - logLik(modelB.coxph)  ), 
       df=3, 
       lower.tail=F) # nope

# We thus conclude that “employment” belongs in the model if “ageGroup4” is also  included, since the p-value for the former is extremely small


```

```{r}

# ANOVA
anova(modelA.coxph, modelC.coxph)
```
```{r AIC}

AIC(modelA.coxph)
AIC(modelB.coxph)
AIC(modelC.coxph) # <- this one

```

### Step AIC
```{r}

modelAll.coxph <- coxph(Surv(ttr, relapse) ~ grp + gender + race + employment + yearsSmoking + levelSmoking +  ageGroup4 + priorAttempts + longestNoSmoke,data=pharmacoSmoking)  

result.step <- step(modelAll.coxph, scope=list(upper=~ grp +  gender + race + employment + yearsSmoking +  levelSmoking + ageGroup4 + priorAttempts +  longestNoSmoke, lower=~grp),data=pharmacoSmoking)  

result.step

ggforest(model = result.step,data = pharmacoSmoking)

```

## 6.5 Smoothing

> A classical treatment of splines is de Boor  [13], and their use in statistics has been discussed by many authors. In survival  analysis, an effective method of finding a smoothing spline is via “penalized partial  likelihood.”

```{r}

modelS4.coxph <- coxph(Surv(ttr, relapse) ~ grp + employment +  + pspline(age, df=4),data = pharmacoSmoking)  
modelS4.coxph 

termplot(modelS4.coxph, se=T, terms=3, ylabs="Log hazard")


```

## Exercises

### 6.1 

> 6.1. The data set “hepatocelluar” is in the “asaur” package. 

It contains 17 clinical  and biomarker measurements on 227 patients, as well as overall survival and  time to recurrence, both recorded in months [42, 43]. There are three measures of CXCL17 activity, CXCL17T (intratumoral), CXCL17P (peritumoral), and  CXCL17N (nontumoral). There is a particular interest in whether they are related to  overall and also recurrence-free survival. 

Which of the three is most strongly related  for each survival outcome? 

For the one most strongly related with survival, fit a  spline model and plot it, as we did in Sect. 6.5. Does this suggest that categorizing  CXCL17 would be appropriate?  


```{r}

attach(hepatoCellular)
data(hepatoCellular)

# hepatoCellular$OS
# hepatoCellular$Death

# hepatoCellular$RFS
# hepatoCellular$Recurrence

# hepatoCellular$CXCL17T
# hepatoCellular$CXCL17P
# hepatoCellular$CXCL17N

## OS

model.CXCL17T.coxph <- coxph(Surv(OS, Death) ~ CXCL17T,data=hepatoCellular)  
model.CXCL17T.coxph
model.CXCL17P.coxph <- coxph(Surv(OS, Death) ~ CXCL17P,data=hepatoCellular)  
model.CXCL17P.coxph
model.CXCL17N.coxph <- coxph(Surv(OS, Death) ~ CXCL17N,data=hepatoCellular)  
model.CXCL17N.coxph

model.CXCL17.coxph <- coxph(Surv(OS, Death) ~ CXCL17T + CXCL17P + CXCL17N ,data=hepatoCellular)  
model.CXCL17.coxph

model.CXCL17.coxph

termplot(model.CXCL17.coxph, se=T, terms=c(1,2,3), ylabs="Log hazard")

```


> 6.2. For the covariates with complete data (in Columns 1–22), 

use stepwise  regression with AIC to identify the best model for (a) overall survival, and  (b) recurrence-free survival. 

```{r}

model.all.coxph = coxph(Surv(OS, Death) ~ .,
                        data=hepatoCellular[,c(2:15,20:22)])  
model.all.coxph.step <- step(model.all.coxph,data=hepatoCellular[,c(2:15,20:22)])  
model.all.coxph.step

```










